'
Data Mining Principles | Assignment 2
Azizha Zeinita | 12334537
'

```{r}
library(ggplot2) 
library(MASS)
library(poLCA)

BostonData <- Boston
```

### --- Step 1 --- ###
# 1.1. Split samples into two random samples of sizes 70% and 30%.
```{r}
crim=BostonData[,1]
rm=BostonData[,6]
age=BostonData[,7]
dis=BostonData[,8]
lstat=BostonData[,13]
medv=BostonData[,14]
df <- data.frame(crim,rm,age,dis,lstat,medv)

X<- sample(c(rep(0, 0.7 * nrow(df)), rep(1, 0.3 * nrow(df))))
table(X)
train <- df[X == 0, ]  
test <- df[X== 1, ] 
```

### --- Step 2 --- ###
# 2.1. Standardize your data so that each variable has mean = 0 and variance = 1
# 2.2. Scale the test set using the mean and standard deviation of the training set. 
```{r}
X.train.mean = colMeans(train)
X.train.sd   = sapply(train, sd)
X.train.scale  = scale(train, center=X.train.mean, scale=X.train.sd)
X.test.scale   = scale(test, center=X.train.mean, scale=X.train.sd) #scaling test by train parameters
```

# 3. Perform PCA on the train data. Use princomp (R) and PCA (Python) function.
```{r}
pc.train <- prcomp(X.train.scale)
```

### --- Step 3 --- ###
# 3.1. Display cumulative sum of variance accounted for (cumulative proportion) by each additional PCA factor.
```{r}
plot (pc.train, type='l')
summary(pc.train)
```

# 3.2. Apply the "elbow rule" to decide how many components you'd like to include.
'Selected number of component: 2'


### --- Step 4 --- ###
# 4.1. Plot loading 1 against all of the other loadings (6 pairwise comparisons). 
#biplot(pc.train[,c(1,2)])
```{r}
biplot(pc.train$x, pc.train$rotation)
```

'Interpretation is documented in PDF document'

# 4.2.1. Show that Component loadings are orthogonal.
```{r}
round(cor(t(pc.train$rotation) %*% pc.train$rotation))
```

# 4.2.2. Show that Component scores are orthogonal.
```{r}
round(cor(t(pc.train$x) %*% pc.train$x))
```

# 4.2.3.1. predict the component scores in the Test using the predict() function in R and transform function in Python
```{r}
pc.test.predict <- predict(pc.train, newdata = X.test.scale)
```

# 4.2.3.2. matrix multiply the predicted component scores from (1) above with transpose of component loadings you derived from training data set from Step 2 above.
```{r}
matrix.multiply <- pc.test.predict[,1:2] %*% t(pc.train$rotation)[1:2,]
```

# 4.2.3.3. Compute the Variance Account For (R2) in the Test sample. That yields a measure of Test performance. 
```{r}
VAF_train <- round(cor(as.vector(X.train.scale), as.vector(pc.train$x[,1:2] %*% t(pc.train$rotation)[1:2,])),2)^2
VAF_test <- round(cor(as.vector(X.test.scale), as.vector(matrix.multiply)),2)^2
VAF_train
VAF_test
```

#Rotating Principal Components Solutions using varimax() in R
```{r}
x=prcomp(df,scale=TRUE)
summary(x)
cor(as.vector(scale(df)), as.vector(x$x[,1:2]  %*% t(x$rotation)[1:2,]))^2
y=varimax(x$rotation[,1:2])
y
cor(as.vector(scale(df)), as.vector(x$x[,1:2] %*% y$rotmat %*% t(y$rotmat) %*% t(x$rotation)[1:2,]))^2
```


